---
title: "Exercise 6"
output: html_document
date: "2025-02-24"
---

For this exercise, you will generate graphs that show univariate, descriptive statistics for nominal, ordinal, and continuous variables.
```{r}
#Use this code chunk to download the NHANES 2017-2018 database and to label the nominal and ordinal variables as factor variables (if necessary):
NHANESdata <- read.csv("https://raw.githubusercontent.com/malenamalka/ANTH241-SP25/refs/heads/main/data/NHANES2017-28")
factor(c(NHANESdata$Gender, NHANESdata$Race, NHANESdata$HealthIns))
factor(c(NHANESdata$Education,NHANESdata$HHEducation,NHANESdata$HHIncome,NHANESdata$FeelBad,NHANESdata$HHFoodSec),order=TRUE)
factor(NHANESdata$BMIcat, order =TRUE, levels = c("Underweight","Healthy","Overweight","Obese"))
head(NHANESdata)
```

1. Load the following packages to complete this exercise: ggplot2, tidyr, and dplyr.
```{r}
library(ggplot2)
library(tidyr)
library(dplyr)
```

1. Bar charts illustrate variation for nominal and ordinal variables.
```{r}
#a. Bar chart for a nominal variable showing raw counts: Create the bar chart for BMI categories (BMIcat)
#ggplot (data, aes(x=Catvar)) + geom_bar() + labs(title="", x="", y="")
NHANESdata %>% 
  drop_na(BMIcat) %>% 
ggplot(aes(x=BMIcat)) +geom_bar(fill="goldenrod") +labs(title="", x="", y="")
```

```{r}
#b. Bar chart showing BMI categories by percentages:
#new_table_name <- with(NHANESdata, table(BMIcat))
#turn the above table into proportions: prop.table(new_table_name)
#turn the table into percentages: prop.table(new_table_name)*100
#new_data_frame <- as.data.frame(prop.table(new_table_name)*100)
#ggplot(new_data_frame, aes(BMIcat, Freq)) + geom_col()

BMIcattable <- with(NHANESdata, table(BMIcat))
prop.table(BMIcattable)
prop.table(BMIcattable)*100
BMIcatframe <- as.data.frame(prop.table(BMIcattable)*100)
ggplot(BMIcatframe, aes(BMIcat, Freq) ) +geom_col(fill="maroon")
```

c. Construct bar charts for one ordinal variable in the code chunk below:
```{r}
NHANESdata %>% 
  drop_na(Education) %>% 
  filter(Education<7) %>% 
  ggplot(aes(x=Education)) + geom_bar(color="mediumpurple3", fill="skyblue") + labs(title= "Highest Level of Education")

NHANESdata %>% 
  drop_na(HHEducation) %>% 
  filter(HHEducation<7) %>% 
  ggplot(aes(x=HHEducation)) + geom_bar(color="peru", fill="navajowhite") + labs(title= "Highest Level of Education")
```

d. Constructing bar charts for a nominal/ordinal variable grouped by another nominal/ordinal variable:
```{r}
#BMI_cats by gender: stacked, grouped and filled bar charts:
#grouping variable= independent
#variation= dependent variable
#table(data$Independent variable, data$Dependent variable)
#Stacked bar chart:
#ggplot(data, aes(x=Independent variable, fill=Dependent variable)) + geom_bar(position = "stack") + labs(title="", x="", y="")


table(NHANESdata$Gender, NHANESdata$BMIcat)
NHANESdata %>% 
  drop_na(BMIcat) %>% 
ggplot(aes(x=BMIcat, fill=Gender)) + geom_bar(position="stack") + labs(title="Stacked Bar Chart", x="BMI Category", y="Gender")

#grouped bar chart:
#ggplot(data, aes(x=Independent variable, fill=dependent variable)) + geom_bar(position = "dodge") + labs(title="", x="Gender", y="Frequency")
NHANESdata %>% 
  drop_na(BMIcat) %>%
  ggplot(aes(x=BMIcat, fill=Gender)) + geom_bar(position="dodge", color="ivory1") +labs(title="Grouped Bar Chart", x="Gender", y="Frequency")

NHANESdata %>% 
  drop_na(BMIcat) %>%
  ggplot(aes(x=BMIcat, fill=Gender)) + geom_bar(position="dodge", color="ivory1") + facet_wrap(~Race) + labs(title="BMI by Gender by Race(Facet Wrap)", x="Gender", y="Frequency") + theme(axis.text.x = element_text(angle=45, hjust=1, size=10))

NHANESdata %>% 
  drop_na(BMIcat) %>%
  ggplot(aes(x=BMIcat, fill=Gender)) + geom_bar(position="dodge", color="ivory1") + facet_grid(~Race) + labs(title="BMI by Gender by Race(Facet Grid)", x="Gender", y="Frequency") + theme(axis.text.x = element_text(angle=45, hjust=1, size=10))

```

e. Constructing bar charts for mean +/- standard deviation of continuous variables
```{r}
#1 generate means and standard errors of BMI(dependent var.) by gender (independent var)
#new_data_frame <- data%>%
#drop_na(Dependent variable)%>%
#group_by(Independent variable)
#summarize(n=n(), mean=mean(dependent variable), sd=sd(dependent variable))
#to print new data frame: new_data_frame

BMIgender <- NHANESdata %>% 
  drop_na(BMI) %>% 
  group_by(Gender) %>% 
  summarize(n=n(), mean=mean(BMI), sd=sd(BMI))
BMIgender

#2 Plot means in a sorted bar chart and adds error bars that represent the standard deviation
#ggplot(new_data_frame, aes(x=reorder(Independent variable, mean), y=mean)) + geom_bar(stat="identity", fill="color") + geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=0.2)
#labs(x="", y="", title="", subtitle="")
ggplot(BMIgender, aes(x=reorder(Gender, mean), y=mean)) + geom_bar(stat="identity", color="floralwhite", fill="mistyrose") + geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=0.3, color="red3") + labs(x="Gender", y="Mean BMI", title="Mean BMI by Gender")
```

2. Histograms illustrate variation in continuous variables:
```{r}
#a. simple histogram: ggplot(data, aes(x=contvar)) + geom_histogram() + labs(title="", x="", y="")
#construct a histogram for BMI:
NHANESdata %>% 
  drop_na(BMI) %>% 
  ggplot(aes(x=BMI)) +geom_histogram(color="floralwhite", fill="springgreen3") + labs(title="BMI", x="BMI", y="")

```

```{r}
#b. colored histogram with 20 bins: ggplot(data, aes(x=contvar)) + geom_histogram(bins=20, color="white", fill="steelblue") + labs(title="Colored histogram with 20 bins", x="", y="")
#do the above code for BMI:
NHANESdata %>% 
  drop_na(BMI) %>% 
  ggplot(aes(x=BMI)) + geom_histogram(bins=30, color="plum4", fill="lavender") +labs(title="Colored Histogram with 30 bins", x="BMI", y="count")

```

```{r}
#c. Histogram with percentages:
#ggplot(data, aes(x=contvar, y=..density..)) + geom_histogram(bins=20, color="white", fill="steelblue") + scale_y_continuous(labels=scales::percent) + labs(title="", y="percent", x="")

NHANESdata %>% 
  drop_na(BMI) %>% 
  ggplot(aes(x=BMI, y=after_stat(density))) + geom_histogram(bins=30, color="slateblue3", fill="powderblue") + scale_y_continuous(labels=scales::percent) +labs(title="BMI", x="BMI", y="Sample Percentage")

NHANESdata %>% 
  drop_na(BMI) %>% 
ggplot(aes(x=BMI, y=..density..)) + geom_histogram(bins=30, color="papayawhip", fill="powderblue")+ labs(x="BMI", y= "Sample Percentage", title="BMI")
```


d: Create a histogram of a different continuous variable:

```{r}
NHANESdata %>% 
  drop_na(WaistCircumf) %>% 
  ggplot(aes(x=WaistCircumf, y=after_stat(density))) + geom_histogram(bins=25, fill="cornflowerblue", color="azure2")
```


3. Box plots: a box and whisker plot describes the distribution of a continuous variable by plotting its five-number summary: minimum, lower quartile (25th percentile), median (50th percentile), upper quartile (75th percentile), and maximum, and the outliers (values outside of the +/- 1.5xQR)
```{r}
#a.ggplot(data, aes(x="", y=Contvar)) + geom_boxplot() + labs(y="", x="", title="")
#Create a box plot of the amount of protein consumed in a day (DR1TPROT)
NHANESdata %>% 
  drop_na(DR1TPROT) %>% 
  ggplot(aes(x="", y=DR1TPROT)) +geom_boxplot(color="lightsalmon2", fill="papayawhip") + labs(y="Grams Protein", x="", title="1-day Protein Intake")

```

(Our box-and-whisker is compressed because we have a LOT of outliers)
         _
         |          <- acceptable data above 75th percentile
|-75th Percentile_|
|-----Median------| <- 50th percentile
|-25th Percentile-|
         |    <- acceptable data below 25th percentile
         -
         o <--- outlier (>1.5x interquartile range)
```{r}
#b. Create a box plot that comnpares a continuous variable across different levels of a nominal or ordinal variable: DR1TPROT by race:
#ggplot(data, aes(x=Independent variable, y=Contvar)) + geom_boxplot(fill="color") + labs(x="Independent variable", y="Contvar", title="")
NHANESdata %>% 
  drop_na(DR1TPROT) %>% 
ggplot(aes(x=Race, y=DR1TPROT)) + geom_boxplot(color="indianred2", fill="floralwhite") + labs(x="Race", y="Daily Protein Intake(Grams)")


NHANESdata %>% 
  drop_na(DR1TPROT) %>% 
ggplot(aes(x=Race, y=DR1TPROT)) + geom_boxplot(color="lightgoldenrod", fill="olivedrab") + facet_wrap(~Gender) + labs(x="Race", y="Daily Protein Intake(Grams)")

NHANESdata %>% 
  drop_na(DR1TPROT) %>% 
ggplot(aes(x=Gender, y=DR1TPROT)) + geom_boxplot(color="slateblue4", fill="slategray2") + facet_wrap(~Race) + labs(x="Gender", y="Daily Protein Intake(Grams)")

```
